Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-05-15 paper/main.tex
deleted my comments in the article

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5's stream in the last 5~Gyr, which -- we remind the reader -- is the time interval investigated in our simulations.  Fig.~\ref{fig:histogram_impact_time} and Table~\ref{tab:gaps_per_perturber} present results of this analysis. NGC~2808 impacted Palomar~5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition} in each case. Since this interaction occurred less than one orbital period ago, despite the uncertainties, the orbital solutions remain similar and thus produce consistent results across all simulations. However, as we continue to turn back time further, the uncertainties in the initial conditions allow the various orbital solutions to diverge from one another. Thus, in one configuration, a cluster can impact the stream at a given time, and yet at the same moment, in a different set of initial conditions, it could be on the other side of the galaxy. \textbf{\salvatore{We will discuss the necessary conditions for creating a gap in Appendix~\ref{sec:gaps_vs_gcmass}. }}

@@

\textbf{\salvatore{An additional explanation for the absence of gaps in the inner regions of a stream is given in App.~\ref{sec:gaps_vs_gcmass}. Briefly, the eccentricity of an orbit induces tidal shocking at the pericenter passages that cause episodes of increased mass loss where the escaped stars leave with a higher mean velocity and greater velocity dispersion. In essence, streams from progenitors on eccentric orbits are made of two components: a continuous flow of stars plus many packets of stars that burst out from pericenter passages. The gap impact occurs at a single position, creating a gap in each subpopulation. However, the gaps have different drift rates across the groups, and they eventually go out of phase, erasing the impact's signature. The packets of stars disperse with time. However, bursts contribute to more escaped stars as the simulation evolves than the continuous outflow. As a result, at late stages of the simulation, the region nearest to the globular cluster is made of distinct yet overlapping populations. After an impact in this location, the impact site between the groups quickly goes out of phase. }}

@@

  \textbf{ \salvatore{\citet{2017MNRAS.470...60E} is another work that discusses the impact of globular clusters in producing gaps on the Pal~5 stream. In section~6.3 of their article, \citet{2017MNRAS.470...60E} discuss the fact that globular clusters with masses greater than $10^6~\textrm{M}_\odot$ are rare in the Galaxy, especially at distances compatible with the orbit of Pal~5, while -- based on previous works -- they estimate that sub-halos of dark matter of similar mass are at least three times greater in number. This difference leads them to conclude that the gaps they report in the Pal~5 streams are more likely to be induced by dark matter sub-halos than by globular clusters. 
  However, as we show in this paper, even clusters with masses below $10^6~\textrm{M}\odot$ can produce gaps. The clusters perturbing the Pal~5 stream are clusters with masses up to 10 times lower than those considered by \citet{2017MNRAS.470...60E}. It is thus possible that \citet{2017MNRAS.470...60E} have underestimated the impact of globular clusters' close passages on a stream such as Pal~5. We can derive from \citet{2016MNRAS.457.3817S}'s third figure that a more concentrated system delivers a higher velocity kick but over a shorter distance, while a more spread out system will affect more stars yet perturb them less. It is more difficult, however, to make a comparative analysis between the role of clusters and sub-halos at equal stellar mass. Being less dense than globular clusters, sub-halos should produce less intense perturbations on streams (for fixed impact parameters and relative velocities). A systematic comparative analysis to quantify the role of clusters and sub-halos in producing gaps and perturbations in stellar streams still needs to be done.}}

@@

  \salvatore{Fig.~\ref{fig:decoherence_shocked_outflow} dissects the more massive stream of Fig.~\ref{fig:gap_no_more} and instructs us why the gap does not persist. The bottom panel shows the total density distribution, and dissects it into two profiles: stars that either are \textit{leaked} or \textit{shocked} out of the cluster. By shocked, we mean the stars that are lost during pericenter passages. Tidal forces generally scale with $r^{-3}$, thus significantly strengthening through their pericenter passages. See Fig.~3 of \citet{2004AJ....127.2753D} for an illustration of the magnitude of the tidal forces along an orbit. A consequence of the pericenter passages is that many more stars escape, fleeing with a higher mean velocity and velocity dispersion, which can be visualized in the \textit{Christmas tree} illustration in Fig.~A.3 of \citet{2012A&A...546L...7M}. We consider the continous outflow of stars between pericenter passages to be \textit{leaked} stars. We identify pericenter passages by computing each local minimum of the cluster's galactic-centric 3D radius with \texttt{scipy.signal.find\_peaks}. The duration of each shock is: $t_{\textrm{shock}}$ = 2$\sqrt{r/|\nabla\Phi|}$, where $r$ is the galactocentric radius and $|\nabla\Phi|$ is the magnitude of the acceleration from the galactic potential. Stars within this range about a given pericenter passage are all given the same color and shown as individual groups in the first, second, third, and fourth panels of Fig.~\ref{fig:decoherence_shocked_outflow}.}

  \salvatore{Each group can be simplified as one-dimensional collisionless dispersion. These packets of stars can be described as a population with the same initial position that drifts about a mean velocity, $\langle v \rangle$, with velocity dispersion, $\sigma_v$. The density evolution is a Gaussian whose amplitude decreases inversely with time, and whose standard deviation grows linearly with time:}
  \begin{equation}    
          \rho(x,t) = \frac{1}{\sigma_v t \sqrt{2\pi} }\textrm{exp}\left(-\frac{1}{2}\left(\frac{x-\langle v \rangle t}{\sigma_v t}\right)^2\right).
  \end{equation}

  \salvatore{It is important to note that within a single packet, the stars self-segregate. The position of the density peak is at $\langle v \rangle t$, while the faster stars are ahead and the slower ones are behind. The packets are separated by a distance of $\langle  v\rangle T$, where $T$ is the orbital period. Regarding our simulations, since the mass and radius of the host are held constant, the mean velocity and velocity dispersion are the same across all packets. Secondly, the density of each packet decreases since the easier-to-remove stars are stripped early on. }

  \salvatore{Consider an impact that happens at $x\prime$. The distance from the impact site to each packet is: $\delta x= x\prime - \langle v\rangle \left(t-n T\right)$, where $n$ is the packet index, with $n=0$ being the first packet escaped at the first pericenter passage in the simulation. Note that $\delta x\propto n$, and velocity segregation means $v\propto x$. Therefore, the different parts of the separate packets are hit at a given impact site. The earlier packets are hit in regions where the stars have lower velocities than the later packets. As a result, the gaps in the later packets quickly overtake those from the earlier packets. As a result, the gaps across the packets go out of phase and interfere destructively.}

  \salvatore{In the simplest consideration, of two gaps of the same dimension $\sigma_x$ and an offset of $\delta x$, the gaps are out of phase when the separation is greater than the width: $1<\sigma_x/\delta x$. In the temporal analysis of the stream presented in Fig.~\ref{fig:decoherence_shocked_outflow}, we did not observe at any point of the simulation that the gap initially persisted to then go out of phase. Drawing from the wisdom in \citet{2015MNRAS.450.1136E} \& \citet{2016MNRAS.457.3817S}, we see the contrast between the density peaks and the density minimum of a gap, as well as the width, grow in time, somewhere between $\propto \sqrt{t}$ and linear. In this simplified 1D collisionless dispersion analysis, the gaps between the different packets space apart linearly in time with $\delta x\propto \delta vt$, where $\delta v$ is the mean velocity of the about the impact site for two different packets. This inclines us to believe that the gaps space apart before they become significantly deep or wide, however this need not be the general case. }

  \salvatore{Nonetheless, we believe this is the first consideration of decomposing the stream into packets of tidal bursts to understand the effect of the internal dynamics on gap persistence. Since gap persistence depends on the mass and internal evolution of the cluster, characterizing the gap can constrain our knowledge of the internal evolution. }
  % \paola{What happens is that the different and in particular in the populations of stars in this part of the tail that escaped from the cluster at different times, in each of these populations the respective gap tends to travel along the streams at different speeds, which leads to a phase shift of the gap and its loss of visibility. In other words, the gap can survive for billions of years in each population of stars that left the cluster at different times, but it is no longer. Still, initially coherent and therefore visible because each of these populations travels along the tail (and drags a gap with it) at different speeds. The loss of coherence occurs at a time $t_{visibility}$ which is directly proportional to the gap size and inversely proportional to the velocity dispersion in that part of the stream. So, if populations of stars lost at different times co-exist in the portion of the stream where the impact occurs, the gap will remain visible for a time that depends on the mass of the progenitor system at the time these stars escaped. From this, we can understand that various factors determine the visibility of a gap in a stream. Mass affects the velocity dispersion in a given region of the stream, but this velocity dispersion, in turn, depends on the history of mass loss of the cluster. Suppose stars dominate a given stream region, and all are lost simultaneously from the cluster. In that case, the resulting velocity dispersion is lower than if various populations coexist in the same region, which has implications for these structures' persistence. }

>>>>

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5's stream in the last 5~Gyr, which -- we remind the reader -- is the time interval investigated in our simulations.  Fig.~\ref{fig:histogram_impact_time} and Table~\ref{tab:gaps_per_perturber} present results of this analysis. NGC~2808 impacted Palomar~5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition} in each case. Since this interaction occurred less than one orbital period ago, despite the uncertainties, the orbital solutions remain similar and thus produce consistent results across all simulations. However, as we continue to turn back time further, the uncertainties in the initial conditions allow the various orbital solutions to diverge from one another. Thus, in one configuration, a cluster can impact the stream at a given time, and yet at the same moment, in a different set of initial conditions, it could be on the other side of the galaxy. We will discuss the necessary conditions for creating a gap in Appendix~\ref{sec:gaps_vs_gcmass}.

@@

An additional explanation for the absence of gaps in the inner regions of a stream is given in App.~\ref{sec:gaps_vs_gcmass}. Briefly, the eccentricity of an orbit induces tidal shocking at the pericenter passages that cause episodes of increased mass loss where the escaped stars leave with a higher mean velocity and greater velocity dispersion. In essence, streams from progenitors on eccentric orbits are made of two components: a continuous flow of stars plus many packets of stars that burst out from pericenter passages. The gap impact occurs at a single position, creating a gap in each subpopulation. However, the gaps have different drift rates across the groups, and they eventually go out of phase, erasing the impact's signature. The packets of stars disperse with time. However, bursts contribute to more escaped stars as the simulation evolves than the continuous outflow. As a result, at late stages of the simulation, the region nearest to the globular cluster is made of distinct yet overlapping populations. After an impact in this location, the impact site between the groups quickly goes out of phase.

@@

  \citet{2017MNRAS.470...60E} is another work that discusses the impact of globular clusters in producing gaps on the Pal~5 stream. In section~6.3 of their article, \citet{2017MNRAS.470...60E} discuss the fact that globular clusters with masses greater than $10^6~\textrm{M}_\odot$ are rare in the Galaxy, especially at distances compatible with the orbit of Pal~5, while -- based on previous works -- they estimate that sub-halos of dark matter of similar mass are at least three times greater in number. This difference leads them to conclude that the gaps they report in the Pal~5 streams are more likely to be induced by dark matter sub-halos than by globular clusters. 
  However, as we show in this paper, even clusters with masses below $10^6~\textrm{M}\odot$ can produce gaps. The clusters perturbing the Pal~5 stream are clusters with masses up to 10 times lower than those considered by \citet{2017MNRAS.470...60E}. It is thus possible that \citet{2017MNRAS.470...60E} have underestimated the impact of globular clusters' close passages on a stream such as Pal~5. We can derive from \citet{2016MNRAS.457.3817S}'s third figure that a more concentrated system delivers a higher velocity kick but over a shorter distance, while a more spread out system will affect more stars yet perturb them less. It is more difficult, however, to make a comparative analysis between the role of clusters and sub-halos at equal stellar mass. Being less dense than globular clusters, sub-halos should produce less intense perturbations on streams (for fixed impact parameters and relative velocities). A systematic comparative analysis to quantify the role of clusters and sub-halos in producing gaps and perturbations in stellar streams still needs to be done.

@@

  Fig.~\ref{fig:decoherence_shocked_outflow} dissects the more massive stream of Fig.~\ref{fig:gap_no_more} and instructs us why the gap does not persist. The bottom panel shows the total density distribution, and dissects it into two profiles: stars that either are \textit{leaked} or \textit{shocked} out of the cluster. By shocked, we mean the stars that are lost during pericenter passages. Tidal forces generally scale with $r^{-3}$, thus significantly strengthening through their pericenter passages. See Fig.~3 of \citet{2004AJ....127.2753D} for an illustration of the magnitude of the tidal forces along an orbit. A consequence of the pericenter passages is that many more stars escape, fleeing with a higher mean velocity and velocity dispersion, which can be visualized in the \textit{Christmas tree} illustration in Fig.~A.3 of \citet{2012A&A...546L...7M}. We consider the continous outflow of stars between pericenter passages to be \textit{leaked} stars. We identify pericenter passages by computing each local minimum of the cluster's galactic-centric 3D radius with \texttt{scipy.signal.find\_peaks}. The duration of each shock is: $t_{\textrm{shock}}$ = 2$\sqrt{r/|\nabla\Phi|}$, where $r$ is the galactocentric radius and $|\nabla\Phi|$ is the magnitude of the acceleration from the galactic potential. Stars within this range about a given pericenter passage are all given the same color and shown as individual groups in the first, second, third, and fourth panels of Fig.~\ref{fig:decoherence_shocked_outflow}.

  Each group can be simplified as one-dimensional collisionless dispersion. These packets of stars can be described as a population with the same initial position that drifts about a mean velocity, $\langle v \rangle$, with velocity dispersion, $\sigma_v$. The density evolution is a Gaussian whose amplitude decreases inversely with time, and whose standard deviation grows linearly with time:
  \begin{equation}    
          \rho(x,t) = \frac{1}{\sigma_v t \sqrt{2\pi} }\textrm{exp}\left(-\frac{1}{2}\left(\frac{x-\langle v \rangle t}{\sigma_v t}\right)^2\right).
  \end{equation}

  It is important to note that within a single packet, the stars self-segregate. The position of the density peak is at $\langle v \rangle t$, while the faster stars are ahead and the slower ones are behind. The packets are separated by a distance of $\langle  v\rangle T$, where $T$ is the orbital period. Regarding our simulations, since the mass and radius of the host are held constant, the mean velocity and velocity dispersion are the same across all packets. Secondly, the density of each packet decreases since the easier-to-remove stars are stripped early on.

  Consider an impact that happens at $x\prime$. The distance from the impact site to each packet is: $\delta x= x\prime - \langle v\rangle \left(t-n T\right)$, where $n$ is the packet index, with $n=0$ being the first packet escaped at the first pericenter passage in the simulation. Note that $\delta x\propto n$, and velocity segregation means $v\propto x$. Therefore, the different parts of the separate packets are hit at a given impact site. The earlier packets are hit in regions where the stars have lower velocities than the later packets. As a result, the gaps in the later packets quickly overtake those from the earlier packets. As a result, the gaps across the packets go out of phase and interfere destructively.

  In the simplest consideration, of two gaps of the same dimension $\sigma_x$ and an offset of $\delta x$, the gaps are out of phase when the separation is greater than the width: $1<\sigma_x/\delta x$. In the temporal analysis of the stream presented in Fig.~\ref{fig:decoherence_shocked_outflow}, we did not observe at any point of the simulation that the gap initially persisted to then go out of phase. Drawing from the wisdom in \citet{2015MNRAS.450.1136E} \& \citet{2016MNRAS.457.3817S}, we see the contrast between the density peaks and the density minimum of a gap, as well as the width, grow in time, somewhere between $\propto \sqrt{t}$ and linear. In this simplified 1D collisionless dispersion analysis, the gaps between the different packets space apart linearly in time with $\delta x\propto \delta vt$, where $\delta v$ is the mean velocity of the about the impact site for two different packets. This inclines us to believe that the gaps space apart before they become significantly deep or wide, however this need not be the general case.

  Nonetheless, we believe this is the first consideration of decomposing the stream into packets of tidal bursts to understand the effect of the internal dynamics on gap persistence. Since gap persistence depends on the mass and internal evolution of the cluster, characterizing the gap can constrain our knowledge of the internal evolution.
